---
title: "2b_data_struct"
output: 
  html_notebook:
    toc: true
    toc_float: true
---

# ***data objects***
data are stored as objects.

## 1. **vector**

### 1.0 Intro
1. one dim array: a systematic arrangement of similar objects
1. vector is the simplest. a scalor is a vector with len=1
1. One vector single data type

### 1.1 create vector

(a). basic  

`v = c(0, 1, 2)`  

(b). other functions

(1) `seq():`  
params: seq(from = 1, to = 1, by = ((to - from)/(length.out - 1))  
note:  
  i start and end are both included
  
### 1.2 accessing elements

```
x[1]
x[1:3]
x[c(1, 3, 5)]
x[x>3]    # all elements that > 3 in x

x[-1]   # exclude first
x[c(-1, -2)]
x[c(-1, -2, 3)]   # results in error
x[-6]   # even if length of x is 5, no error
```
  
### 1.3 calculation

1. directly use math symbols:

```
x=5:1

x > 2   # a vector of TRUE and FALSE
x * 2   # elewise, same with other op
```

1. diff(v)  
return the diff `v[i+1]-v[i]`  
```
> diff(c(1, 2, 4))
[1] 1 2
```

### 1.4 misc operations

1. `:`  
    Similar to the basics of `seq`  
    
    ```
    1:5 gets 1 2 3 4 5
    5:1 gets 5 4 3 2 1
    ```

1. length()  

1. combination of vectors  
    use `c()`. Can combine diff datatype, but will convert to the same:  
        1. numeric + logical: logical F -> 0, T -> 1
        1. character + numerical/logical: numerical/logical -> char

1. explicit conversion of data type
    `as.numeric(v)`: eg characters that cannot be converted will be `NA`

1. unique()  
    return unique elements as a list  
    
    ```
    > unique(c("C", "D", "H", "D", "H"))
    [1] "C" "D" "H"
    ```

1. split()  
    split the vector to several vectors and return as a list  
    
    ```
    > g = c('a', 'b', 'c', 'a', 'b', 'c', 'a', 'b', 'c')
    > x = c(1, 2, 3, 4, 5,  6, 7, 8, 9)
    > split(x, g)
    $a
    [1] 1 4 7
    
    $b
    [1] 2 5 8
    
    $c
    [1] 3 6 9
    ```
    
1. repeat  

    ```
    # repeat a number: rep(x, times)
    rep(3, 12)
    
    
    # repeat a vector:
    rep(seq(1,3), 2)        # 1, 2, 3, 1, 2, 3 
    rep(c(1,2,3), c(1,2,3)) # 1, 2, 2, 3, 3, 3; note: if num not match, error
    rep(c(1,2,3), each=2)   # 1, 1, 2, 2, 3, 3
    
    
    rep(seq(1,3), each=2)        # 1, 2, 3, 1, 2, 3 
    ```
    
1. `tapply`
    1. Intro: Apply a function to subsets of a vector or array, split by factors
    
    1. Syntax
        `tapply(data, INDEX, FUN, ...)`

        data: The vector or array you want to summarize.  
        INDEX: A list of factors or grouping variables used to split the data.  
        FUN: The function you want to apply to each subset.  
    1. Eg.  
    
        ```
        scores <- c(85, 90, 78, 92, 88, 76, 84, 92, 95, 89)
        grades <- c("A", "A", "B", "A", "B", "C", "B", "A", "A", "B")

        # Using tapply() to calculate the average score for each grade
        tapply(scores, grades, mean)
        
            A     B     C 
        90.80 84.75 76.00 
        ```
      


## 2. **matrix (& multi dimensional arrays)**

### 0. intro

n by m means n rows m cols

### 1. creation
`matrix(data = NA, nrow = 1, ncol = 1, byrow = FALSE,dimnames = NULL)`
```
# col-wise fill in by default
> m<-matrix(1:12,nrow=3,ncol=4)
> m
     [,1] [,2] [,3] [,4]
[1,]    1    4    7   10
[2,]    2    5    8   11
[3,]    3    6    9   12

matrix(1:9, nrow=3) # no error
matrix(1:9, nrow=3) # no error
matrix(1:10, nrow=3) # error

```
### 2. inspection
```
dim(m)
nrow(m)
ncol(m)
```

### 3. access

```
m[2,3]    # element, slicing, -1 ... same with vector
m[2,]     # select row (result is vector)
m[,3]     # select col (result is vector)

m[1:2, 3:4]   # (result is a matrix)

```
### 4. misc op

1. combine:  

      ```
      m = cbind(m1, m2)   # bind in axis of col. If m1 is axb, m2 is cxb, then m is (a+c)xb
      m = rbind(m1, m2)   # similarly
      
      ```

1. transpose: `t()`

1. `prop.table(rs,margin=1)`: along axis=margin (1 for row, 2 for col), return the proportion (percentage)

1. `col()`, `row()`  
    Returns a matrix of integers indicating their row number in a matrix-like object, or a factor indicating the row labels.
    
    ```
    > row(matrix(0, 2, 3))
         [,1] [,2] [,3]
    [1,]    1    1    1
    [2,]    2    2    2
    ```
  

### z. side track: mul-dim array

1. create  
    `array(data = NA, dim = length(data), dimnames = NULL)`  
    mind the dim! first 2 dim is the array, then is the third dim  
    also mind for filling  
    
    ```
    > array(data=1:24, dim=c(4, 3, 2))
    , , 1
    
         [,1] [,2] [,3]
    [1,]    1    5    9
    [2,]    2    6   10
    [3,]    3    7   11
    [4,]    4    8   12
    
    , , 2
    
         [,1] [,2] [,3]
    [1,]   13   17   21
    [2,]   14   18   22
    [3,]   15   19   23
    [4,]   16   20   24
    
    ```

## 3. **list**

### 3.0. intro
most general obj. compared to vectors and matrices: can hold diff type of obj

### 3.1. creation
```
> w<-list(1:3,c("a","b"),T)
[[1]]
[1] 1 2 3

[[2]]
[1] "a" "b"

[[3]]
[1] TRUE
```

### 3.2. name

```
assign name: 
> names(w) = c("x", "y", "z")
$x
[1] 1 2 3

$y
[1] "a" "b"

$z
[1] TRUE
```

### 3.3. access
2 subscript. first (double [] denot component of the list, single []: ele in the component)  

1. names are access with $; index are accessed with [[]]

```
w[[2]]  # OK
w$x     # OK
w$"x"   # OK

w[["x"]]    # error
w$1         # error
```

1. special case 1: access name with `[]`
```
> w["x"]
$x
[1] 1 2 3
```

1. special case 2: partially write out the name
```
names(w) = c("ab" ,"y", "z")

# using `$`, can partially
> w$a
[1] 1 2 3

# using `[]`, cannot
> w["a"]
$<NA>
NULL
```


### 3.4 misc op

1. `unlist()`: change the list obj to a long vector, using similar data type casting rule


## 4. **dataframe**

### 4.1. intro

data frame is a special kind of list where each member (col, but not row) are vectors of equal length  

```
> women   # built-in dataframe in R
   height weight
1      58    115
2      59    117
3      60    120
4      61    123
5      62    122
```

Each row is called an **observation** and each column is called a **variable**  

### 4.2 create dataframe
```
member <- data.frame( 
  name = c("Tom", "May"), 
  age = c(22, 20) 
)
```

### 2.3 access
Can only access columns first as it is a "list"  

1. Similar to list, `[[]]` or `$`: return as a vector
```
> women$weight
 [1] 115 117 120 123 126 129 132 135 139 142 146 150 154 159 164
> women[[2]]
 [1] 115 117 120 123 126 129 132 135 139 142 146 150 154 159 164
```

1. `with()`: directly access columns w/o using $:
```
with(women, weight/height)
# gives out a vector: [1] 1.982759 1.983051 2.000000 2.016393 ...
```

### 2.4 vector-wise operations

1. `by()`: 
    similar to `tapply()`, but the return obj class is "by"
    1. Syntax:  
        `by(data, INDICES, FUN)`

### 2.5 common untrivial operations
1. Find the elements that satisfy some rule

    ```
    > women$height[women$weight > 140]    
    [1] 67 68 69 70 71 72  
    ```
    
    explain: 
    `women$weight > 140` is vector operation and result is a vector of T and F  
     then use this T F vector to find elements

### 2.4 misc methods

1. `summary(df)`: provides the mean max ...

## 5. **factor**

### Intro

Is effective when there are a lot of same elements: categorize the data, represent & store it on multiple levels.  
A level is the representation of a kind of element. Name of levels are string. Levels coded (stored) as integers (so effective to store data with repeated element).

### create

```{r}
# create from list
> grp <- factor(c("control", "treatment", "control", "treatment"))
> grp
[1] control   treatment control   treatment
Levels: control treatment
```

### inspect
```{r}
# see the code (int representation) of levels
> as.integer(grp) 
[1] 1 2 1 2

# inspect levels
> levels(grp) 
[1] "control" "treatment"

> levels(grp)[as.integer(grp)] 
[1] "control" "treatment" "control" "treatment"
```

### operations
```{r}
# relabel
> levels(grp) <- c("1","2") 
> grp 
[1] 1 2 1 2 
Levels: 1 2

# compare / find wanted element
# Note: need to compare as string
> grp == "1" 
[1]  TRUE FALSE  TRUE FALSE
```


