---
title: "basic concept and syntax"
output: html_notebook
---

# ***basic built-in operations***

## 1. **system and R lang related**

1. help: `help(<method_name>)`
1. `mode()`: shows the "data type"  
    common data type: numeric, logical (TRUE and FALSE)


# ***data objects***
data are stored as objects.

## 1. **vector**

### 1.0 Intro
1. one dim array: a systematic arrangement of similar objects
1. vector is the simplest. a scalor is a vector with len=1
1. One vector single data type

### 1.1 create vector

(a). basic  

`v = c(0, 1, 2)`  

(b). other functions

(1) `seq():`  
params: seq(from = 1, to = 1, by = ((to - from)/(length.out - 1))  
note:  
  i start and end are both included
  
### 1.2 accessing elements

```
x[1]
x[1:3]
x[c(1, 3, 5)]
x[x>3]    # all elements that > 3 in x

x[-1]   # exclude first
x[c(-1, -2)]
x[c(-1, -2, 3)]   # results in error
x[-6]   # even if length of x is 5, no error
```
  
### 1.3 calculation

1. directly use math symbols:

```
x=5:1

x > 2   # a vector of TRUE and FALSE
x * 2   # elewise, same with other op
```

1. diff(v)  
return the diff `v[i+1]-v[i]`  
```
> diff(c(1, 2, 4))
[1] 1 2
```




### 1.4 misc operations

1. length()  

1. combination of vectors  
use `c()`. Can combine diff datatype, but will convert to the same:  
    1. numeric + logical: logical F -> 0, T -> 1
    1. character + numerical/logical: numerical/logical -> char

1. explicit conversion of data type
`as.numeric(v)`: eg characters that cannot be converted will be `NA`

1. unique()  
return unique elements as a list  
```
> unique(c("C", "D", "H", "D", "H"))
[1] "C" "D" "H"
```

### 1.5 related misc function

a. `:`
```
1:5 gets 1 2 3 4 5
5:1 gets 5 4 3 2 1
```
a. repeat

```{r}
# repeat a number: rep(x, times)
rep(3, 12)


# repeat a vector:
rep(seq(1,3), 2)        # 1, 2, 3, 1, 2, 3 
rep(c(1,2,3), c(1,2,3)) # 1, 2, 2, 3, 3, 3; note: if num not match, error
rep(c(1,2,3), each=2)   # 1, 1, 2, 2, 3, 3


rep(seq(1,3), each=2)        # 1, 2, 3, 1, 2, 3 
```

## 2. **matrix (& multi dimensional arrays)**

### 0. intro

n by m means n rows m cols

### 1. creation
`matrix(data = NA, nrow = 1, ncol = 1, byrow = FALSE,dimnames = NULL)`
```
# col-wise fill in by default
> m<-matrix(1:12,nrow=3,ncol=4)
> m
     [,1] [,2] [,3] [,4]
[1,]    1    4    7   10
[2,]    2    5    8   11
[3,]    3    6    9   12

matrix(1:9, nrow=3) # no error
matrix(1:9, nrow=3) # no error
matrix(1:10, nrow=3) # error

```
### 2. inspection
```
dim(m)
nrow(m)
ncol(m)
```

### 3. access

```
m[2,3]    # element, slicing, -1 ... same with vector
m[2,]     # select row (result is vector)
m[,3]     # select col (result is vector)

m[1:2, 3:4]   # (result is a matrix)

```
### 4. misc op

1. combine:
```
m = cbind(m1, m2)   # bind in axis of col. If m1 is axb, m2 is cxb, then m is (a+c)xb
m = rbind(m1, m2)   # similarly

```

### mul-dim array

1. create
`array(data = NA, dim = length(data), dimnames = NULL)`  
mind the dim! first 2 dim is the array, then is the third dim  
also mind for filling  

```
> array(data=1:24, dim=c(4, 3, 2))
, , 1

     [,1] [,2] [,3]
[1,]    1    5    9
[2,]    2    6   10
[3,]    3    7   11
[4,]    4    8   12

, , 2

     [,1] [,2] [,3]
[1,]   13   17   21
[2,]   14   18   22
[3,]   15   19   23
[4,]   16   20   24

```

## 3. **list**

### 3.0. intro
most general obj. compared to vectors and matrices: can hold diff type of obj

### 3.1. creation
```
> w<-list(1:3,c("a","b"),T)
[[1]]
[1] 1 2 3

[[2]]
[1] "a" "b"

[[3]]
[1] TRUE
```

### 3.2. name

```
assign name: 
> names(w) = c("x", "y", "z")
$x
[1] 1 2 3

$y
[1] "a" "b"

$z
[1] TRUE
```

### 3.3. access
2 subscript. first (double [] denot component of the list, single []: ele in the component)  

1. names are access with $; index are accessed with [[]]

```
w[[2]]  # OK
w$x     # OK
w$"x"   # OK

w[["x"]]    # error
w$1         # error
```

1. special case 1: access name with `[]`
```
> w["x"]
$x
[1] 1 2 3
```

1. special case 2: partially write out the name
```
names(w) = c("ab" ,"y", "z")

# using `$`, can partially
> w$a
[1] 1 2 3

# using `[]`, cannot
> w["a"]
$<NA>
NULL
```


### 3.4 misc op

1. `unlist()`: change the list obj to a long vector, using similar data type casting rule


## 2. **dataframe**

### 2.1. intro

data frame is a special kind of list where each member (col, but not row) are vectors of equal length  

```
> women   # built-in dataframe in R
   height weight
1      58    115
2      59    117
3      60    120
4      61    123
5      62    122
```

Each row is called an **observation** and each column is called a **variable**  

### 2.2 create dataframe
```{r}
member <- data.frame( 
  name = c("Tom", "May"), 
  age = c(22, 20) 
)
```

### 2.3 access
Can only access columns first as it is a "list"  

1. Similar to list, `[[]]` or `$`: return as a vector
```
> women$weight
 [1] 115 117 120 123 126 129 132 135 139 142 146 150 154 159 164
> women[[2]]
 [1] 115 117 120 123 126 129 132 135 139 142 146 150 154 159 164
```

1. `with()`: directly access columns w/o using $:
```
with(women, weight/height)
# gives out a vector: [1] 1.982759 1.983051 2.000000 2.016393 ...
```

### 2.5 common untrivial operations
```
# ~
women$height[women$weight > 140]    # [1] 67 68 69 70 71 72  
## explain: 
##`women$weight > 140` is vector operation and result is a vector of T and F
## then use this T F vector to find elements
```

### 2.4 misc methods

1. `summary(df)`: provides the mean max ...


# ***common built-in operations***

## 1. **random**

### 1.1. random sample

(a). `sample()`

1. syntax  
`sample(x, size, replace = FALSE, prob = NULL)`
`sample.int(n, size = n, replace = FALSE, prob = NULL, useHash = (n > 1e+07 && !replace && is.null(prob) && size <= n/2))`

2. usage  
```{r}
# basics
sample(c(1, 2, 3), size=2)
sample(52, size = 5, replace=F)   # means sample from 1:52

# without size param: means sample same num of (permutation)
sample(10)  # permutation of 1:10 (since replace=F)

# sample with probability
sample(c(-1, 0, 1), size=2, prob=c(0.25, 0.5, 0.25))
```

## 2. **sorting**

1. sort
`sort(x, decreasing = FALSE, ...)`
```
> sort(c(3, 12, 2, 1))
[1]  1  2  3 12
```

1. give sort order
```
# 2 methods both OK
> order(c(3, 12, 2, 1))
[1] 4 3 1 2

> sort.list(c(3, 12, 2, 1))
[1] 4 3 1 2
```





# ***misc***

1. polyroot()  

1. imaginary number
    1. Im()
    
1. all



# ***tricky***

1. ```y<-x<2``` is actually: ```y = (x<2)```
